---
title: "Statistical Inference Course Project - Analysis of *tooth growth* data"
author: "Jakob Schelbert"
date: "2.12.2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(datasets)
library(dplyr)
library(ggplot2)
library(knitr)
set.seed(1337)
```


```{r dataloading, echo=FALSE}
# loading the data
toothdata <- ToothGrowth
```

# Synopsis
This report analyzes the **tooth growth** data which contains data obtained from measuring the tooth growth of guinea pigs.
Two different supplements each with three different doses are included in the data set.


# Data description
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC). (taken from help page of *ToothGrowth*)

[,1]	 len	 numeric	 Tooth length
[,2]	 supp	 factor	 Supplement type (VC or OJ).
[,3]	 dose	 numeric	 Dose in milligrams/day



```{r expl_analysis}
toothsum <- summary(toothdata)
toothdata %>% group_by(dose, supp) %>% summarize(n=n(), min=min(len), mean = mean(len), median = median(len), max=max(len), sd = sd(len))
g <- ggplot(data = toothdata, aes(x = supp, y = len))
g + geom_boxplot() + geom_jitter(width = 0.1) + facet_grid(. ~ dose)

g <- ggplot(data = toothdata, aes(x = len))
g + geom_histogram(bins = 5) + facet_grid(dose ~ supp)

```





# Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
